---
title: "MM2-eksamen"
author: "Mads Hovaldt"
date: "28/10/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(gRbase)
library(gRain)
```

We have the dataset ChestSim1000
```{r}
data(chestSim1000, package="gRbase")
head(chestSim1000)
```

This is a hyphotetical Chest Clinic problem, by Lauritzen and Spiegelhalter. (ref til https://arxiv.org/pdf/1301.7394.pdf)

Here is a short explanation of the variables in the dataset.

* asia $\rightarrow$ subject has visited asia
* tub $\rightarrow$ subject has tuberculosis
* smoke $\rightarrow$ subject is a smoker
* lung $\rightarrow$ subject has lung cancer
* bronc $\rightarrow$ subject has bronchitis
* either $\rightarrow$ subject has either tuberculosis or lungcancer
* xray $\rightarrow$ subject has positive X-ray
* dysp $\rightarrow$ Subject has dyspnoea

Shortness-of-breath (dyspnoea) may be due to tuberculosis, lung cancer, bronchitis, none of them, or more than one of them. A recent visit to Asia increases the chances of tuberculosis, while smoking is known to be a risk factor for both lung cancer and bronchitis. The results of a single chest X-ray do not discriminate between lung cancer and tuberculosis, as does neither the presence nor absence of dyspnoea. (citat direkte sat ind fra https://arxiv.org/pdf/1301.7394.pdf)

```{r}
dg1 <- dag(~ S + L|S + X|L:S + B|S + D|L:B)
plot(dg1)
```




```{r}
dg2<-dag(~ asia+tub|asia+ either|tub:lung+smoke+lung|smoke+bronc|smoke+dysp|either:bronc+xray|either)
plot(dg2)
```


# Exercise 1 
Extract the necessary CPT's from data, and construct the Bayesian network.

# Answer

```{r} 
Counting1 <- function(x,y,z){ #counting the number of observation for x and y
 if (x!=0 && y!=0 && z==0){
  a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes")))
  a12 <- (length(which(chestSim1000[x] == "no" & chestSim1000[y] == "yes")))
  a21 <- (length(which(chestSim1000[x] == "yes" & chestSim1000[y] == "no")))
  a22 <- (length(which(chestSim1000[x] == "no" & chestSim1000[y] == "no")))
  mat1 <- matrix(c(a11,a12, a21,a22), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c("yes","no")))
  mat2 <- names(dimnames(mat1))<-c(y,x)
  print(mat1)
  } else if(y==0 && z==0) { # counting the number of observations given x and y=0
  a11 <- (length(which(chestSim1000[x]  == "yes")))
  a12 <- (length(which(chestSim1000[x] == "no")))
  mat1 <- matrix(c(a11,a12),1,2 , dimnames = list(c(" "),c("yes","no")))
  mat2 <- names(dimnames(mat1))<-c(" ",x)
  print(mat1)
  } else if(x!=0 && y!=0 && z!=0){ # counting the number of observations given x,y and z
    a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes")))
    a12 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes")))
    a13 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes")))
    a14 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes")))
    a21 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no")))
    a22 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no")))
    a23 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no")))
    a24 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no")))
    mat1 <- matrix(c(NA,a11,a12,a13,a14, NA,a21,a22,a23,a24), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c(x,"yes/ yes","no/yes","yes/no","no/no")))
    mat2<-names(dimnames(mat1))<-list(z,y)
    print(mat1)
  }
}
Counting1("asia","tub","0")
Counting1("lung","tub","either") #why this result?
Counting1("lung","tub","0") # no one has both lung and tub 
Counting1("bronc","either","dysp")
# Now we create a list of our CPT
library(gRain)
asia1 <- cptable(~asia, values = Counting1("asia","0","0"), levels = c("yes","no"))
tub.asia1<-cptable(~tub|asia, values = Counting1("asia","tub","0"),levels = c("yes","no"))
smoke1<-cptable(~smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))
lung.smoke1<-cptable(~lung|smoke,values = Counting1("smoke","lung","0"),levels = c("yes","no"))
bronc.smoke1<-cptable(~bronc|smoke,values = Counting1("smoke","bronc","0"),levels = c("yes","no"))
either.lung.tub1<-cptable(~either|lung:tub,values = Counting1("lung","tub","either")[,2:5],levels = c("yes","no")) 
xray.either1<-cptable(~xray|either,values = Counting1("either","xray","0"),levels = c("yes","no"))
dysp.bronc.either1<-cptable(~dysp|bronc:either,values = Counting1("bronc","either","dysp")[,2:5],levels = c("yes","no"))
CTP.List1 <- compileCPT(list(asia1,tub.asia1,smoke1,lung.smoke1,bronc.smoke1,either.lung.tub1,xray.either1,dysp.bronc.either1))
CTP.List1
CTP.List1$asia
CTP.List1$tub
CTP.List1$smoke
CTP.List1$lung
CTP.List1$bronc
CTP.List1$either # the normal method can't read the CPT when we have 3 variables
CTP.List1$xray
CTP.List1$dysp # the normal method can't read the CPT when we have 3 variables
ftable(CTP.List1$either, row.vars = 1)
ftable(CTP.List1$dysp, row.vars = 1)
```

# Exercise 2
What does information about "dysp" tell us about "smoke", i.e. what is the conditional distribution of "smoke" given "dysp"?

# Answer
```{r}
dysp1 <- cptable(~ dysp,values = Counting1("dysp","0","0"),levels = c("yes","no"))
smoke.dysp1 <- cptable(~smoke|dysp,values = Counting1("dysp","smoke","0"),levels = c("yes","no"))
CTP.list2 <- compileCPT(list(dysp1,smoke.dysp1))
CTP.list2
CTP.list2$smoke
```

# Exercise 3
If we know "smoke", what does additional infomration about "bronc" tell us about "lung"? That is, what is the conditional distribution of "lung" given smoke", and what is the conditional distribution of "lung" given "smoke" and "bronc"?

# Answer
```{r}
smoke1 <- cptable(~ smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))
bronc1 <- cptable(~bronc,values = Counting1("bronc","0","0"),levels = c("yes","no"))
lung.smoke1 <- cptable(~lung1|smoke,values = Counting1("smoke","lung","0"),levels = c("yes","no"))
lung.smoke.bronc1 <- cptable(~lung2|smoke:bronc,values = Counting1("smoke","bronc","lung")[,2:5],levels = c("yes","no"))
CTP.list3 <- compileCPT(list(smoke1,bronc1,lung.smoke1,lung.smoke.bronc1))
CTP.list3
CTP.list3$lung1
CTP.list3$lung2
ftable(CTP.list3$lung2, row.vars = 1)
```

# Exercise 4
If we know "smoke" and "dysp", what does additional information about "bronc" tell us about "lung"?

# Answer
```{r}
Counting2 <- function(x,y,z,w){ #counting the number of observation for x,y,z and w
    a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a12 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a13 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a14 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a15 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a16 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a17 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a18 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    
    a21 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a22 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a23 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a24 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a25 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a26 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a27 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a28 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    mat1 <- matrix(c(a11,a12,a13,a14,a15,a16,a17,a18, a21,a22,a23,a24,a25,a26,a27,a28), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c("yyy","nyy","yny","nny","yyn","nyn","ynn","nnn")))
    mat2<-names(dimnames(mat1))<-list(w,c(x,y,z))
    print(mat1)
}
Counting2("smoke","dysp","bronc","lung")
Counting1("smoke", "dysp", "lung")
smoke1 <- cptable(~ smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))
dysp1 <- cptable(~ dysp,values = Counting1("dysp","0","0"),levels = c("yes","no"))
bronc1 <- cptable(~ bronc,values = Counting1("bronc","0","0"),levels = c("yes","no"))
lung.smoke.dysp1 <- cptable(~lung1|smoke:dysp, values = Counting1("smoke", "dysp", "lung")[,2:5],levels = c("yes","no"))
lung.smoke.dysp.bronc1 <- cptable(~lung2|bronc:dysp:smoke, values = Counting2("bronc","dysp","smoke","lung"),levels = c("yes", "no"))
CTP.list4 <- compileCPT(list(smoke1,dysp1,bronc1,lung.smoke.dysp1,lung.smoke.dysp.bronc1))
CTP.list4
CTP.list4$lung1
ftable(CTP.list4$lung1, row.vars = 1)
CTP.list4$lung2
ftable(CTP.list4$lung2, row.vars = 1)
```

# Exercise 5
Sketch the message passing algorithm for finding clique marginals for this specific example

# Answer
```{r}
Bay.net1 <- grain(CTP.List1)
Bay.net1 <- compile(Bay.net1)
Bay.net1
plot(Bay.net1$dag)
plot(moralize(Bay.net1$dag))
plot(triangulate(moralize(Bay.net1$dag)))
Bay.net2 <- grain(CTP.list2)
Bay.net2 <- compile(Bay.net2)
Bay.net2
plot(Bay.net2$dag)
plot(moralize(Bay.net2$dag))
plot(triangulate(moralize(Bay.net2$dag)))
Bay.net3 <- grain(CTP.list3)
Bay.net3 <- compile(Bay.net3)
Bay.net3
plot(Bay.net3$dag)
plot(moralize(Bay.net3$dag))
plot(triangulate(moralize(Bay.net3$dag)))
Bay.net4 <- grain(CTP.list4)
Bay.net4 <- compile(Bay.net4)
Bay.net4
plot(Bay.net4$dag)
plot(moralize(Bay.net4$dag))
plot(triangulate(moralize(Bay.net4$dag)))
```

# Part 2, initial work
Consider the "cad" data in "gRbase". There are two dataset: "cad1" which is complete and "cad2" which has missing values here and there.
```{r}
library(graph)
library(Rgraphviz)
library(RBGL)
library(gRbase)
library(gRain)
library(bnlearn)
library(magrittr)
data(cad1, package="gRbase")
names(cad1)

use <- c("Sex", "CAD", "Inherit", "Smoker", "Hyperchol", "Heartfail", "AMI")
dat1 <- cad1[, use]
```

# Exercise 6
use the hill climbing algorithm function from the "bnlearn" package to estimate different Bayesian networks based on data "cad1". see R script provided elsewhere.

# Answer
```{r}
## Start search from empty graph
mm1 <- hc(dat1)
mm1

## Start search from complete graph
sat <-random.graph(use, prob = 1)
mm2 <- hc(dat1, start=sat)
mm2

par(mfrow=c(1,2))
plot(mm1)
plot(mm2)

## Create Bayesian network
bn1 <- as.grain(bn.fit(mm1, dat1))
bn2 <- as.grain(bn.fit(mm2, dat1))
```

# Exercise 7
Predict the value of the CAD variable in the dataset "cad1" for each of the models you find. Predict the value of the CAD variable in the dataset "cad2" for each of the models you find. Is it most appropriate to evaluate the models based on "cad1" or "cad2".

# Answer
```{r}
## Predict data

## Sample 40 random rows
set.seed(1213)
userow <- sample(nrow(dat1), 40)
wdat1 <- dat1[userow,]

pred1 <- predict(bn1, newdata=wdat1, response="CAD")
pred2 <- predict(bn2, newdata=wdat1, response="CAD")

table(wdat1$CAD, pred1$pred$CAD)
table(wdat1$CAD, pred2$pred$CAD)

## Procent 

table(wdat1$CAD, pred1$pred$CAD)/40*100
table(wdat1$CAD, pred2$pred$CAD)/40*100

#' Notice:
#' 
#' Prediction based on same data as we used for fitting / model search
#' is cheating. Use cad2 data instead.
#'
#' What are the misclassification errors under various models?
#'
#' Which misclassifications are the most serious ones?
#' 
#' Using cad2:

data(cad2, package="gRbase")
names(cad2)

use <- c("Sex", "CAD", "Inherit", "Smoker", "Hyperchol", "Heartfail", "AMI")
dat2 <- cad2[, use]

## Sample 40 random rows
set.seed(1213)
userow <- sample(nrow(dat2), 40)
wdat2 <- dat2[userow,]

pred3 <- predict(bn1, newdata=wdat2, response="CAD")
pred4 <- predict(bn2, newdata=wdat2, response="CAD")

table(wdat2$CAD, pred3$pred$CAD)
table(wdat2$CAD, pred4$pred$CAD)
```

# Exercise 8
Compute the misclassification probabilities for persons with CAD and persons without CAD for each model. Which misclassification is most severe?

# Answer
```{r}
## Procenter
table(wdat2$CAD, pred3$pred$CAD)/40*100
table(wdat2$CAD, pred4$pred$CAD)/40*100

```

Type 1 error: to reject, while the patient is positive

Type 2 error: to accept, while the patient is negative

It can be seen that a type 1 and type 2 error is the same for "pred4" for cad2"
 
For "pred3" for "cad2" there is 15% type 1 error and 10% type 2 error

If the treatment is hard on the patients, a type 2 error would be problematic, however introducing multiple test would avoid this

Type 1 errors are "the most servere" since the patient wouldn't get the treatment

this can lead to large consequences for the patient.




