---
title: "MM2-eksamen"
author: "Mads Hovaldt"
date: "28/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(gRbase)
library(gRain)
```

We have the dataset ChestSim1000
```{r}
data(chestSim1000, package="gRbase")
head(chestSim1000)
```

This is a hyphotetical Chest Clinic problem, by Lauritzen and Spiegelhalter. (ref til https://arxiv.org/pdf/1301.7394.pdf)

Here is a short explanation of the variables in the dataset.

* asia $\rightarrow$ subject has visited asia
* tub $\rightarrow$ subject has tuberculosis
* smoke $\rightarrow$ subject is a smoker
* lung $\rightarrow$ subject has lung cancer
* bronc $\rightarrow$ subject has bronchitis
* either $\rightarrow$ subject has either tuberculosis or lungcancer
* xray $\rightarrow$ subject has positive X-ray
* dysp $\rightarrow$ Subject has dyspnoea

Shortness-of-breath (dyspnoea) may be due to tuberculosis, lung cancer, bronchitis, none of them, or more than one of them. A recent visit to Asia increases the chances of tuberculosis, while smoking is known to be a risk factor for both lung cancer and bronchitis. The results of a single chest X-ray do not discriminate between lung cancer and tuberculosis, as does neither the presence nor absence of dyspnoea. (citat direkte sat ind fra https://arxiv.org/pdf/1301.7394.pdf)

```{r}
dg1 <- dag(~ S + L|S + X|L:S + B|S + D|L:B)
plot(dg1)
```




```{r}
dg2<-dag(~ asia+tub|asia+ either|tub:lung+smoke+lung|smoke+bronc|smoke+dysp|either:bronc+xray|either)
plot(dg2)
```


# Opgave 1 
```{r} 


Counting1 <- function(x,y,z){ #counting the number of observation for x and y
 if (x!=0 && y!=0 && z==0){
  a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes")))
  a12 <- (length(which(chestSim1000[x] == "no" & chestSim1000[y] == "yes")))
  a21 <- (length(which(chestSim1000[x] == "yes" & chestSim1000[y] == "no")))
  a22 <- (length(which(chestSim1000[x] == "no" & chestSim1000[y] == "no")))
  mat1 <- matrix(c(a11,a12, a21,a22), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c("yes","no")))
  mat2 <- names(dimnames(mat1))<-c(y,x)
  print(mat1)
  } else if(y==0 && z==0) { # counting the number of observations given x and y=0
  a11 <- (length(which(chestSim1000[x]  == "yes")))
  a12 <- (length(which(chestSim1000[x] == "no")))
  mat1 <- matrix(c(a11,a12),1,2 , dimnames = list(c(" "),c("yes","no")))
  mat2 <- names(dimnames(mat1))<-c(" ",x)
  print(mat1)
  } else if(x!=0 && y!=0 && z!=0){ # counting the number of observations given x,y and z
    a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes")))
    a12 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes")))
    a13 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes")))
    a14 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes")))
    a21 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no")))
    a22 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no")))
    a23 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no")))
    a24 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no")))
    mat1 <- matrix(c(NA,a11,a12,a13,a14, NA,a21,a22,a23,a24), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c(x,"yes/ yes","no/yes","yes/no","no/no")))
    mat2<-names(dimnames(mat1))<-list(z,y)
    print(mat1)
  }
}
Counting1("asia","tub","0")
Counting1("lung","tub","either") #why this result?
Counting1("lung","tub","0") # no one has both lung and tub 
Counting1("bronc","either","dysp")

# Now we create a list of our CPT
library(gRain)
asia1 <- cptable(~asia, values = Counting1("asia","0","0"), levels = c("yes","no"))

tub.asia1<-cptable(~tub|asia, values = Counting1("asia","tub","0"),levels = c("yes","no"))

smoke1<-cptable(~smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))

lung.smoke1<-cptable(~lung|smoke,values = Counting1("smoke","lung","0"),levels = c("yes","no"))

bronc.smoke1<-cptable(~bronc|smoke,values = Counting1("smoke","bronc","0"),levels = c("yes","no"))

either.lung.tub1<-cptable(~either|lung:tub,values = Counting1("lung","tub","either")[,2:5],levels = c("yes","no")) 

xray.either1<-cptable(~xray|either,values = Counting1("either","xray","0"),levels = c("yes","no"))

dysp.bronc.either1<-cptable(~dysp|bronc:either,values = Counting1("bronc","either","dysp")[,2:5],levels = c("yes","no"))

CTP.List1 <- compileCPT(list(asia1,tub.asia1,smoke1,lung.smoke1,bronc.smoke1,either.lung.tub1,xray.either1,dysp.bronc.either1))
CTP.List1

CTP.List1$asia
CTP.List1$tub
CTP.List1$smoke
CTP.List1$lung
CTP.List1$bronc
CTP.List1$either # the normal method can't read the CPT when we have 3 variables
CTP.List1$xray
CTP.List1$dysp # the normal method can't read the CPT when we have 3 variables
ftable(CTP.List1$either, row.vars = 1)
ftable(CTP.List1$dysp, row.vars = 1)

```

# Opgave 2
```{r}
dysp1 <- cptable(~ dysp,values = Counting1("dysp","0","0"),levels = c("yes","no"))
smoke.dysp1 <- cptable(~smoke|dysp,values = Counting1("dysp","smoke","0"),levels = c("yes","no"))

CTP.list2 <- compileCPT(list(dysp1,smoke.dysp1))
CTP.list2
CTP.list2$smoke
```

# Opgave 3
```{r}
smoke1 <- cptable(~ smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))
bronc1 <- cptable(~bronc,values = Counting1("bronc","0","0"),levels = c("yes","no"))
lung.smoke1 <- cptable(~lung1|smoke,values = Counting1("smoke","lung","0"),levels = c("yes","no"))
lung.smoke.bronc1 <- cptable(~lung2|smoke:bronc,values = Counting1("smoke","bronc","lung")[,2:5],levels = c("yes","no"))

CTP.list3 <- compileCPT(list(smoke1,bronc1,lung.smoke1,lung.smoke.bronc1))
CTP.list3
CTP.list3$lung1
CTP.list3$lung2
ftable(CTP.list3$lung2, row.vars = 1)
```

# Opgave 4
```{r}
Counting2 <- function(x,y,z,w){ #counting the number of observation for x and y
    a11 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a12 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a13 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a14 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a15 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a16 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    a17 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "yes")))
    a18 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "yes")))
    
    a21 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a22 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a23 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a24 <- (length(which(chestSim1000[x]  == "yes" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a25 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a26 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "yes" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    a27 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "yes" & chestSim1000[w] == "no")))
    a28 <- (length(which(chestSim1000[x]  == "no" & chestSim1000[y] == "no" & chestSim1000[z] == "no" & chestSim1000[w] == "no")))
    mat1 <- matrix(c(a11,a12,a13,a14,a15,a16,a17,a18, a21,a22,a23,a24,a25,a26,a27,a28), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c("yyy","nyy","yny","nny","yyn","nyn","ynn","nnn")))
    mat2<-names(dimnames(mat1))<-list(w,c(x,y,z))
    print(mat1)
}
Counting2("smoke","dysp","bronc","lung")
Counting1("smoke", "dysp", "lung")

smoke1 <- cptable(~ smoke,values = Counting1("smoke","0","0"),levels = c("yes","no"))
dysp1 <- cptable(~ dysp,values = Counting1("dysp","0","0"),levels = c("yes","no"))
bronc1 <- cptable(~ bronc,values = Counting1("bronc","0","0"),levels = c("yes","no"))
lung.smoke.dysp1 <- cptable(~lung1|smoke:dysp, values = Counting1("smoke", "dysp", "lung")[,2:5],levels = c("yes","no"))
lung.smoke.dysp.bronc1 <- cptable(~lung2|bronc:dysp:smoke, values = Counting2("bronc","dysp","smoke","lung"),levels = c("yes", "no"))

CTP.list4 <- compileCPT(list(smoke1,dysp1,bronc1,lung.smoke.dysp1,lung.smoke.dysp.bronc1))
CTP.list4
CTP.list4$lung1
ftable(CTP.list4$lung1, row.vars = 1)
CTP.list4$lung2
ftable(CTP.list4$lung2, row.vars = 1)


```

# Opgave 5
```{r}
Bay.net1 <- grain(CTP.List1)
Bay.net1 <- compile(Bay.net1)
Bay.net1
plot(Bay.net1$dag)
plot(moralize(Bay.net1$dag))
plot(triangulate(moralize(Bay.net1$dag)))

Bay.net2 <- grain(CTP.list2)
Bay.net2 <- compile(Bay.net2)
Bay.net2
plot(Bay.net2$dag)
plot(moralize(Bay.net2$dag))
plot(triangulate(moralize(Bay.net2$dag)))

Bay.net3 <- grain(CTP.list3)
Bay.net3 <- compile(Bay.net3)
Bay.net3
plot(Bay.net3$dag)
plot(moralize(Bay.net3$dag))
plot(triangulate(moralize(Bay.net3$dag)))

Bay.net4 <- grain(CTP.list4)
Bay.net4 <- compile(Bay.net4)
Bay.net4
plot(Bay.net4$dag)
plot(moralize(Bay.net4$dag))
plot(triangulate(moralize(Bay.net4$dag)))
```

















