---
title: "Præsentation"
author: "Anders"
date: "28/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(gRbase)
```

We have the dataset ChestSim1000
```{r}
data(chestSim1000, package="gRbase")
head(chestSim1000)
```

This is a hyphotetical Chest Clinic problem, by Lauritzen and Spiegelhalter. (ref til https://arxiv.org/pdf/1301.7394.pdf)

Here is a short explanation of the variables in the dataset.

* asia $\rightarrow$ subject has visited asia
* tub $\rightarrow$ subject has tuberculosis
* smoke $\rightarrow$ subject is a smoker
* lung $\rightarrow$ subject has lung cancer
* bronc $\rightarrow$ subject has bronchitis
* either $\rightarrow$ subject has either tuberculosis or lungcancer
* xray $\rightarrow$ subject has positive X-ray
* dysp $\rightarrow$ Subject has dyspnoea

Shortness-of-breath (dyspnoea) may be due to tuberculosis, lung cancer, bronchitis, none of them, or more than one of them. A recent visit to Asia increases the chances of tuberculosis, while smoking is known to be a risk factor for both lung cancer and bronchitis. The results of a single chest X-ray do not discriminate between lung cancer and tuberculosis, as does neither the presence nor absence of dyspnoea. (citat direkte sat ind fra https://arxiv.org/pdf/1301.7394.pdf)

```{r}
dg1 <- dag(~ S + L|S + X|L:S + B|S + D|L:B)
plot(dg1)
```



```{r}
P1 <- function(obj, i, j){
  x <- unlist(obj[which(names(obj) == i)])
  y <- unlist(obj[which(names(obj) == j)])
  a11 <- length(which(x == "yes" & y == "yes"))/length(x)
  a12 <- length(which(x == "no" & y == "yes"))/length(x)
  a21 <- length(which(x == "yes" & y == "no"))/length(x)
  a22 <- length(which(x == "no" & y == "no"))/length(x)
  mat <- matrix(c(a11,a12, a21,a22), nrow = 2, byrow = TRUE, dimnames = list(c("yes","no"),c("yes","no")))
  names(dimnames(mat)) <- c(j,i)
  print(mat)
}
P1(chestSim1000, "asia", "tub")

P1(chestSim1000, "smoke", "lung")
```


















